---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(bdftools)
```


```{r eval=FALSE, include=FALSE}
pkgdown::build_site(override = list(destination = "../coolbutuseless.github.io/package/bdftools"))
```

# bdftools


<!-- badges: start -->
![](https://img.shields.io/badge/cool-useless-green.svg)
<!-- badges: end -->

`bdftools` provides some tools for reading, manipulating and outputting BDF bitmap fonts.
    
### What's in the box

* `read_bdf(filename)` reads a BDF pixel font file into an R representation
   i.e. an object of class `bdf`
* `print.bdf()` prints meta information and a font sample. 
* `bdf_create_df(bdf, text)` Create a data.frame of points for the given string 
* `bdf_create_mat(bdf, text)` Create a matrix representation for the given string
* `bdfGrob(bdf, text, ...)` create a simple grob representation of the given string using
   squares for pixels
* `as.data.frame.bdf(bdf)` converts the full `bdf` font into a rectangular 
   data.frame of all characters and their (x, y) coordinates.
* `read_bdf_builtin()` to read in a font included with this package:
    * [Cozette](https://github.com/slavfox/Cozette) License: MIT. See `LICENSE-cozette`
    * [Creep2](https://github.com/raymond-w-ko/creep2) License: MIT. See `LICENSE-creep2`
    * [Spleen](https://github.com/fcambus/spleen) License: BSD 2-clause. See `LICENSE-spleen`

You can install from [GitHub](https://github.com/coolbutuseless/bdftools) with:

``` r
# install.package('remotes')
remotes::install_github('coolbutuseless/bdftools')
```

## Example: Loading and displaying a BDF bitmap font.

```{r comment='', fig.height = 3}
library(grid)
library(ggplot2)
library(bdftools)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Load a BDF font
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
fontfile <- system.file("fonts", "spleen-5x8.bdf", package = "bdftools", mustWork = TRUE)
myfont <- read_bdf(fontfile)

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# By default, printing a font will print some header info, and a text 
# sample rendered in that font in the console
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
myfont

#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
# Generate some sample text in the console
#~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~~
create_matrix("Hello RStats", myfont) |> 
  as.raster() |>
  plot(interpolate = FALSE)
```


```{r comment='', fig.height = 3}
coords <- create_coords("Hello\n#RStats!", myfont)
head(coords)

grid.newpage()
grid.rect(coords$x * 4, coords$y * 4, width = 3, height = 3, default.units = 'mm',
          gp = gpar(fill = 'lightblue'))
```




# An example of a larger font

```{r comment=""}
myfont <- bdftools::read_bdf_builtin("spleen-16x32.bdf")
create_matrix("Frak!", myfont) |> 
  as.raster() |>
  plot(interpolate = FALSE)
```



## Font sample sheet

An example of how the data.frame representation of the font can be plotted 
in `ggplot2`.

```{r fig.width = 8, fig.height = 8}
library(ggplot2)
library(bdftools)

myfont <- read_bdf_builtin("cozette.bdf")

plot_df <- as.data.frame(myfont)
plot_df <- plot_df[plot_df$encoding >= utf8ToInt('A') & plot_df$encoding <= utf8ToInt('|'),]

ggplot(plot_df) +
  geom_tile(aes(x, y), width=0.9, height = 0.9, na.rm = TRUE) +
  facet_wrap(~encoding + desc, ncol = 12)+
  theme_void(10) +
  coord_equal()
```





## BDF Bitmap Font Resources

* Fonts
    * [Tom Thumb tiny font](https://robey.lag.net/2010/01/23/tiny-monospace-font.html) (License: Dual licensed
      CC0 or CC-BY 3.0 license.)
    * [github repo of lots of fonts](https://github.com/Tecate/bitmap-fonts) (License: Various/unknown)
    * [unifont](https://www.unifoundry.com/unifont/index.html) a BDF font with glyphs
      for every printable codepoint from `U+0000` to `U+FFFF`.  (License: Dual licensed 
      SIL Open Font License (OFL) version 1.1 and the GNU GPL 2+ with the GNU font embedding exception)
    * [X11 fonts](https://www.cl.cam.ac.uk/~mgk25/ucs-fonts.html) (License: Public Domain)
* BDF Font Specification
    * [BDF on Wikipedia](https://en.wikipedia.org/wiki/Glyph_Bitmap_Distribution_Format)
    * [BDF Specification Document (Adobe)](https://adobe-type-tools.github.io/font-tech-notes/pdfs/5005.BDF_Spec.pdf)

## Related Software

Other retro/vector/pixel fonts in R

* [arcade font](https://github.com/coolbutuseless/arcadefont)
* [gridfont](https://github.com/coolbutuseless/gridfont)
* [hershey](https://github.com/coolbutuseless/hershey)


