---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = FALSE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)

library(bdftools)
```


```{r eval=FALSE, include=FALSE}
pkgdown::build_site(override = list(destination = "../coolbutuseless.github.io/package/bdftools"))
```

# bdftools


<!-- badges: start -->
![](https://img.shields.io/badge/cool-useless-green.svg)
<!-- badges: end -->

`bdftools` provides some tools for reading, manipulating and outputting BDF bitmap fonts.
    
### What's in the box

* Bitmap fonts
    * `bitmap_text_coords()` returns a data.frame of pixel locations
    * `bitmap_text_matrix()` returns a binary matrix with pixel locations set to 1
    * `bitmap_text_raster()` returns a raster image of the text
* Vector font
    * `vector_text_coords()` returns a data.frame of strokes
    * `vector_text_matrix()` returns a binary matrix with pixel locations set to 1
    * `vector_text_raster()` returns a raster image of the text

### Fonts


| Type   | Name            | Sizes                         | Unicode?              | # glyphs |
|--------|-----------------|-------------------------------|-----------------------|----------|
| Bitmap | Cozette         | 20x12 ??                      | Some                  | 3536     |
| Bitmap | Creep2          | 11x11 ??                      | Some                  | 508      |
| Bitmap | Spleen          | 5x8, 6x12, 8x16, 12x24, 16x32 | Some                  | 450-1000 |
| Bitmap | Unifont         | 16x16                         | Yes!                  | 57086    |
| Vector | gridfont        |                               | Lower ASCII only      |          |
| Vector | gridfont_smooth |                               | Lower ASCII only      |          |
| Vector | arcade          |                               | Upper case ASCII only |          |


Bitmap fonts:

* [Cozette v1.25.2](https://github.com/slavfox/Cozette) License: MIT. See `LICENSE-cozette.txt`
* [Creep2](https://github.com/raymond-w-ko/creep2) License: MIT. See `LICENSE-creep2.txt`
* [Spleen v2.1.0](https://github.com/fcambus/spleen) License: BSD 2-clause. See `LICENSE-spleen.txt`
* [Unifont](https://unifoundry.com/unifont/) License: SIL Open Font License (OFL) version 1.1. See `LICENSE-unifont.txt`

Vector fonts

* [gridfont](https://github.com/inconvergent/gridfont) License: MIT. See `LICENSE-gridfont.txt`
* `arcade` is a vector font I created. License: SIL Open Font License (OFL) version 1.1. See `LICENSE-arcade.txt`


## Installation

You can install from [GitHub](https://github.com/coolbutuseless/bdftools) with:

``` r
# install.package('remotes')
remotes::install_github('coolbutuseless/bdftools')
```

## Example: Loading and displaying a BDF bitmap font.

```{r comment='', fig.height = 3}
library(grid)
library(ggplot2)
library(bdftools)

c("cozette_hidpi", "cozette", "creep2-11", "spleen-12x24", "spleen-16x32", 
"spleen-32x64", "spleen-5x8", "spleen-6x12", "spleen-8x16", "unifont")

txt <- "Hello #RStats"
txt <- "привет"
txt <- "二項分布\xF0\x9F\x8E\xB2の英語表記は\n「Binomial distribution」である。"


bitmap_text_raster(txt, "unifont") |> 
  plot(interpolate = FALSE)
```


```{r comment='', fig.height = 3}
library(grid)
coords <- bitmap_text_coords("Helljq\n#RSta!!!", "unscii-16")
head(coords)

grid.newpage()
grid.rect(coords$x * 4, coords$y * 4, width = 3, height = 3, default.units = 'mm',
          gp = gpar(fill = 'lightblue'))
```




## Bit map font - sample sheet

An example of how the data.frame representation of the font can be plotted 
in `ggplot2`.

```{r fig.width = 8, fig.height = 8}
library(ggplot2)
library(bdftools)

txt <- "привет"
txt <- "二項分布\xF0\x9F\x8E\xB2の英語表記は「Binomial distribution」である。"
# txt <- "Hello"
plot_df <- bitmap_text_coords(paste(txt, collapse = ""), "unifont")
# plot_df <- bitmap_text_coords(paste(txt, collapse = ""), "spleen-32x64")
# plot_df <- bitmap_text_coords(paste(txt, collapse = ""), "unscii-16")

ggplot(plot_df) +
  geom_tile(aes(x0, y0), width=0.9, height = 0.9, na.rm = TRUE) +
  facet_wrap(~idx, ncol = 12)+
  theme_void(10) +
  coord_equal()
```


## Vector Font

```{r}
df <- vector_text_coords('Hello', font = 'gridfont')
df

vector_text_matrix('Hello', font = 'gridfont', scale = 1)

ras <- vector_text_raster('Hello\n#RStats!', font = 'gridfont_smooth', scale = 15)
plot(ras, interpolate = FALSE)
```


## Font Sheets

```{r}
lo <- utf8ToInt('!')
hi <- utf8ToInt('z')
txt <- intToUtf8(seq(lo, hi))

coords <- bitmap_text_coords(txt, "spleen-6x12")

ggplot(coords) +
  geom_tile(aes(x0, y0), width=0.9, height = 0.9, na.rm = TRUE) +
  facet_wrap(~idx, ncol = 12)+
  theme_void(10) +
  coord_equal()
```




## BDF Bitmap Font Resources

* Fonts
    * [Tom Thumb tiny font](https://robey.lag.net/2010/01/23/tiny-monospace-font.html) (License: Dual licensed
      CC0 or CC-BY 3.0 license.)
    * [github repo of lots of fonts](https://github.com/Tecate/bitmap-fonts) (License: Various/unknown)
    * [unifont](https://www.unifoundry.com/unifont/index.html) a BDF font with glyphs
      for every printable codepoint from `U+0000` to `U+FFFF`.  (License: Dual licensed 
      SIL Open Font License (OFL) version 1.1 and the GNU GPL 2+ with the GNU font embedding exception)
    * [X11 fonts](https://www.cl.cam.ac.uk/~mgk25/ucs-fonts.html) (License: Public Domain)
* BDF Font Specification
    * [BDF on Wikipedia](https://en.wikipedia.org/wiki/Glyph_Bitmap_Distribution_Format)
    * [BDF Specification Document (Adobe)](https://adobe-type-tools.github.io/font-tech-notes/pdfs/5005.BDF_Spec.pdf)

## Related Software

Other retro/vector/pixel fonts in R

* [arcade font](https://github.com/coolbutuseless/arcadefont)
* [gridfont](https://github.com/coolbutuseless/gridfont)
* [hershey](https://github.com/coolbutuseless/hershey)


